---
title: "Modeling Rat Growth"
author: "Gilberto Padilla Mercado"
format: html
---

```{r}
library(dplyr)
library(ggplot2)
source("fitting.R")
rat_df <- readr::read_tsv("ratBW_data1.tsv")
```

# Data Cleanup

For these models, we want to ensure that all data is essentially summary data,
with N > 1 and a variance associated. When variance is not included but N > 1,
it will be set to the median variance.

```{r}
maxCV <- sqrt(rat_df$Variance)
maxCV <- maxCV/rat_df$Weight
maxCV <- median(maxCV, na.rm= TRUE)
rat_df <- rat_df |> 
  filter(!is.na(N), !Sex %in% "both", N > 1) |> 
  mutate(Variance = ifelse(is.na(Variance),
                           (maxVarCV * Weight)^2,
                           Variance),
         Studies = as.factor(paste(Species, Reference, Sex, sep = "_")))

ggplot(rat_df, aes(x = Age, y = Weight, color = Sex)) + 
  geom_point(shape = 21) +
  geom_line(aes(group = Studies)) +
  scale_color_manual(values = c("black", "green4")) +
  theme_bw() + 
  theme(legend.position = "bottom")
```

# Nonlinear Regression Model

# Growth Models

We can model the growth using a nonlinear mixed effects model. Three types of models:  

Gompertz:  $W(t) = A \cdot \exp(- \exp(-k \cdot (t-t_0))$  
Logistic:  $W(t) = \frac{A}{1+\cdot \exp(-k \cdot (t-t_0))}$  


For all of these the parameter as as follows:  

-  `A` is the mature body weight.  
-  `t_0` is the time at inflection point.
-  `k` is the rate of growth.  


## Logistic


```{r}
logistic_mod <- list(
  name = "logistic",
  starts = function(x, y) {
    N <- length(unique(x[x < 300]))
    if (N %% 2 == 1) { # ODD
      ix <- 1+(N%/%2)
      ix1 <- ix - 1
      ix2 <- ix + 1
      ek <- mean(c(
        (y[ix]-y[ix1])/(x[ix]-x[ix1]),
        (y[ix2]-y[ix])/(x[ix2]-x[ix])
      ))
    } else { # EVEN
      ix <- 1+(N%/%2) - 1
      ix2 <- ix+1
      ek <- (y[ix2]-y[ix])/(x[ix2]-x[ix])
    }
    c(
      A = 0.4,
      k = (signif(ek, 3)*4),
      tmid = round(median(x[x < 300]))
    )
  },
  lower_lim = c(A = 0.01, k = 0.0001, tmid = 0),
  upper_lim = c(A = 1, k = 1, tmid = 300),
  predict = function(input, A, k, tmid) {
    value <- A / (1 + (exp(-k * (input - tmid))))
    return(value)
  }
)

```

## Gompertz

```{r}
gompertz_mod <- list(
  name = "Gompertz",
  starts = function(x, y) {
    N <- length(unique(x[x < 300]))
    if (N %% 2 == 1) { # ODD
      ix <- 1+(N%/%2)
      ix1 <- ix - 1
      ix2 <- ix + 1
      ek <- mean(c(
        (y[ix]-y[ix1])/(x[ix]-x[ix1]),
        (y[ix2]-y[ix])/(x[ix2]-x[ix])
      ))
    } else { # EVEN
      ix <- 1+(N%/%2) - 1
      ix2 <- ix+1
      ek <- (y[ix2]-y[ix])/(x[ix2]-x[ix])
    }
    c(
      A = 0.4,
      k = (signif(ek, 3)*4),
      tmid = round(median(x[x < 300]))
    )
  },
  lower_lim = c(A = 0.01, k = 0.0001, tmid = 0),
  upper_lim = c(A = 1, k = 1, tmid = 300),
  predict = function(input, A, k, tmid) {
    value <- A * exp(-exp(-k * (input - tmid)))
    return(value)
  }
)
```



```{r}
all_fits <- fit_all_groups(rat_df,
                           grp_fct = "Sex", err_fct = c("Species", "Reference"),
                           mfun = list(logistic_mod, gompertz_mod))

ggplot(rat_df, aes(x = Age, y = Weight, color = Sex)) + 
  geom_point(shape = 21) +
  scale_color_manual(values = c("black", "green4")) +
  geom_line(aes(y = logistic_mod$predict(Age, A=0.235, k=0.04377, tmid=50.29)),
            color = "black") +
  geom_line(aes(y = logistic_mod$predict(Age, A=0.3809, k=0.0441, tmid=59.9)),
            color = "green4") +
  theme_bw() + 
  theme(legend.position = "bottom")
```


