---
title: "Modeling Rat Growth"
author: "Gilberto Padilla Mercado"
format: html
---

```{r}

library(ggplot2)
library(tictoc)
source("fitting.R")
rat_df <- readr::read_tsv("ratBW_data1.tsv")
```

# Data Cleanup

For these models, we want to ensure that all data is essentially summary data,
with N > 1 and a variance associated. When variance is not included but N > 1,
it will be set to the median variance.

```{r}
maxCV <- sqrt(rat_df$Variance)
maxCV <- maxCV/rat_df$Weight
maxCV <- median(maxCV, na.rm= TRUE)
rat_df <- subset(rat_df, 
                 !is.na(N) & !(Sex %in% "both") & N > 1)
rat_df$Studies <- with(
  rat_df, factor(paste(Species, Reference, Sex, sep = "_"))
)

rat_df$Variance <- with(
  rat_df,
  ifelse(is.na(Variance), (maxCV * Weight)^2, Variance)
)

ggplot(rat_df, aes(x = Age, y = Weight, color = Sex)) + 
  geom_point(shape = 21) +
  geom_line(aes(group = Studies)) +
  scale_color_manual(values = c("black", "green4")) +
  theme_bw() + 
  theme(legend.position = "bottom")
```

# Nonlinear Regression Model

# Growth Models

We can model the growth using a nonlinear mixed effects model. Three types of models:  

Gompertz:  $W(t) = A \cdot \exp(- \exp(-k \cdot (t-t_0))$  
Logistic:  $W(t) = \frac{A}{1+\cdot \exp(-k \cdot (t-t_0))}$  


For all of these the parameter as as follows:  

-  `A` is the mature body weight.  
-  `t_0` is the time at inflection point.
-  `k` is the rate of growth.  


## Logistic


```{r}
logistic_mod <- list(
  name = "logistic",
  starts = function(x, y) {
    # tmid has to be the point at which Y = A/2
    # so find closest Y in data to this value based on start.
    A <- 0.4
    tmid <- x[which.min(abs(y - (A/2)))]
    tq <- x[which.min(abs(y - (A/4)))]
    k <- log(4 - 1)/(tmid - tq)
    list(
      A = A,
      k = signif(max(1E-4, k), 4),
      tmid = round(tmid)
    )
  },
  lower_lim = list(A = 0.01, k = 1E-5, tmid = 2),
  upper_lim = list(A = 1, k = 1, tmid = 300),
  predict = function(Age, A, k, tmid) {
    value <- A / (1 + (exp(-k * (Age - tmid))))
    return(value)
  }
)

all_fits <- fit_data(rat_df, 
                     data_fct = "Sex", sigma_fct = "Sex",
                     mod_fun = logistic_mod)

ggplot(rat_df, aes(x = Age, y = Weight, color = Sex)) + 
  geom_point(shape = 21, alpha = 0.4) +
  geom_function(fun = \(x) do.call(logistic_mod$predict,
                                   c(list(Age = x), get_coefs(all_fits$male))),
                color = "green4", linewidth = 1, alpha = 0.8) +
  geom_function(fun = \(x) do.call(logistic_mod$predict,
                                   c(list(Age = x), get_coefs(all_fits$female))),
                color = "black", linewidth = 1, alpha = 0.8) +
  scale_color_manual(values = c("black", "green4")) +
  theme_bw() + 
  theme(legend.position = "bottom")
```

## Gompertz

```{r}
gompertz_mod <- list(
  name = "Gompertz",
  starts = function(x, y) {
    N <- length(unique(x[x < 300]))
    if (N %% 2 == 1) { # ODD
      ix <- 1+(N%/%2)
      ix1 <- ix - 1
      ix2 <- ix + 1
      ek <- mean(c(
        (y[ix]-y[ix1])/(x[ix]-x[ix1]),
        (y[ix2]-y[ix])/(x[ix2]-x[ix])
      ))
    } else { # EVEN
      ix <- 1+(N%/%2) - 1
      ix2 <- ix+1
      ek <- (y[ix2]-y[ix])/(x[ix2]-x[ix])
    }
    c(
      A = 0.4,
      k = (signif(ek, 3)*4),
      tmid = round(median(x[x < 300]))
    )
  },
  lower_lim = c(A = 0.01, k = 0.0001, tmid = 0),
  upper_lim = c(A = 1, k = 1, tmid = 300),
  predict = function(input, A, k, tmid) {
    value <- A * exp(-exp(-k * (input - tmid)))
    return(value)
  }
)
```



```{r}


all_fits <- fit_data(rat_df, data_fct = "Sex", sigma_fct = "Species",
                     mod_fun = logistic_mod)

ggplot(rat_df, aes(x = Age, y = Weight, color = Sex)) + 
  geom_point(shape = 21) +
  scale_color_manual(values = c("black", "green4")) +
  geom_line(aes(y = logistic_mod$predict(Age, A=0.235, k=0.04377, tmid=50.29)),
            color = "black") +
  geom_line(aes(y = logistic_mod$predict(Age, A=0.3809, k=0.0441, tmid=59.9)),
            color = "green4") +
  theme_bw() + 
  theme(legend.position = "bottom")
```


